---
title: "Hackathon Level 1"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    toc_depth: 2
editor_options: 
  markdown: 
    wrap: 72
---


<!-- <style> -->
<!-- .list-group-item.active, .list-group-item.active:focus, .list-group-item.active:hover { -->
<!--     background-color: #08C477; -->
<!-- } -->
<!-- </style> -->


```{=html}
<style type="text/css">
  body{
  font-size: 12pt;
}
</style>
```

```{r include=FALSE}
knitr::opts_chunk$set( eval = TRUE, echo= TRUE, message = FALSE, warning = FALSE, highlight = TRUE)
```

# Introduction

We will analyse 10x Genomics Visium spatial transcriptomics (ST) data from a tissue slide of **high-grade serous ovarian carcinoma** (HGSOC).\

Ovarian cancer remains the eighth leading cause of cancer deaths in women worldwide and high-grade serous ovarian carcinoma (HGSOC) is the most common and lethal histologic subtype. \
Genomes of HGSOC are **highly heterogeneous** with most alterations only found in a small fraction of tumours. \
Also, due to a high degree of **chromosomal instability**, most HGSOCs are polyclonal. As the cancer progresses and metastasises, **clonal diversity** increases, which is associated with **worse prognosis** and development of **chemoresistance**. \

Standard first-line treatment for HGSOC typically consists of debulking surgery, which involves removal of as much of the tumor as possible, followed by intravenous paclitaxel/platinum-based chemotherapy, and often subsequent maintenance therapy.\ 

Investigating tumor heterogeneity, The Cancer Genome Atlas (TCGA) project revealed **four molecular subtypes** of HGSOC based on bulk expression measurements: mesenchymal, proliferative, immunoreactive, and differentiated. These subtypes are associated with differences in prognosis. \
It is now generally accepted that transcriptional subtypes of HGSOC largely reflect the **degree of immune cell infiltration and the abundance of fibroblasts**, rather than inherent differences in tumour cells. \
To determine how these non-malignant cell types might influence tumour growth and prognosis, ST data hold the potential to reconstruct ligand-receptor interactions between stromal, immune and tumour cell populations, resolving the tumor tissue architecture at spatial resolution. \


In this hackathon we will analyze data of tumour sample collected during interval debulking surgery from HGSOC patient with a good response to taxane- and platinum-based neoadjuvant chemotherapy (NACT) treatment.

Data were retrieved from [Denisenko, E. et al., Spatial transcriptomics reveals discrete tumour microenvironments and autocrine loops within ovarian cancer subclones. Nat Commun. (2024)](https://doi.org/10.1038/s41467-024-47271-y)


```{=html}
  <hr>
```


# Warming up: setting the environment that you need

Before starting, let's set the stage to be all on the same ground.\
You have been provided with configuration files. Please select "your" environment and the "helper" functions\

```{r, setting_envir, message=TRUE, highlight=TRUE}
source('../00_environment.R')
source("../00_helper_functions.R")
```

```{=html}
  <hr>
```


# Loading the packages you need

```{r libraries , message=FALSE}
source("level_1_libraries.R")
```

```{=html}
  <hr>
```


# Loading 10x Visium data of HGSOC slide {.tabset .tabset-fade .tabset-pills}

To create the `SeuratObject` representing the ST data of the sequenced slide we will rely 
on information stored in the `input` folder. 
This folder contains `raw counts` and the hematoxylin and eosin (HE) image of the
tissue slide:\

| What       | Path to                                              |
|:-----------|:-----------------------------------------------------|
| raw counts | `../input/10XVisium/`                                |
| HE image   | `../input/10XVisium/spatial/tissue_lowres_image.png` |

```{=html}
  <hr>
```


In `raw counts` folder there are three files usually provided by 10x Genomics Visium:\

-   **matrix.mtx:**\
    Count matrix is usually stored as a sparse format for a more
    efficient manipulation.\
    Rows: genes\
    Columns: barcodes associated to each spot.\
    \
-   **features.tsv:**\
    This file contains specifications of rows included in the count
    matrix file.\
    For each feature, the following information are reported:\

     -   first column: `feature ID`, *e.g* gene ID "ENSG00000284733"
    -   second colum: `feature name`, *e.g* gene name "OR4F29"
    -   third column: `feature type` which will be one of the following
        options: \
          - Gene Expression (this case) \
          - Antibody Capture \
          - CRISPR Guide Capture \
          - Multiplexing Capture \
          - CUSTOM \
\
-   **barcodes.tsv:**\
    This table contains barcode sequences corresponding to column
    indices of the matrix file.\
    Each barcode sequence contains a suffix with a dash separator
    followed by a number, such as '-1'.\


```{=html}
  <hr>
```


## Question {.unnumbered}

Referring to paths reported in the table above, load 10X Visium data in a `SeuratObject`.\
A `SeuratObject` is a specialized data structure developed in the [Seurat package](https://github.com/satijalab/seurat), enabling the study of gene expression in the context
of tissue architecture.\
This `SeuratObject` is a container that organizes and stores both the
spot-level expression data along with the associated image of the tissue
slide.\

Now, create a `SeuratObject` containing spot-level gene expression counts
and the image of tissue slide; in order:\

1.  Read 10X Genomics Visium count matrix using `Read10X()` function\


2.  Create a `SeuratObject` with the count matrix with
    `CreateSeuratObject()` \
    
    
3.  Read 10X Genomics Visium image through `Read10X_Image()`. \
    Be careful to include only spots determined to be over tissue.



```{=html}
  <hr>
```



**Questions**\

A) How many spots do you consider in the assay?\


B) How many genes are included in the panel?\



```{=html}
  <hr>
```


## Answer {.unnumbered}


```{r}
# Your code here















```

```{=html}
  <hr>
```



# Test your dataset {.tabset .tabset-fade .tabset-pills}


The resulting `sp` object contains in `meta.data` table:

-   `orig.ident`: the identifier of the sample\
-   `nCount_spatial`: the total number of counts per spot\
-   `nFeature_spatial`: the number of detected features (genes) per
    spot\



```{=html}
  <hr>
```



## Question {.unnumbered}

Now, you have to:\

1.  Measure the `percentage of MT` genes of each spot and quantify the
    `capture efficiency` per spot.\
    The `capture efficiency` is defined as the ratio of number of
    expressed genes (`nFeature`) and total counts (`nCount`) per spot.\
    Add these two annotations in the meta.data table included in
    your `SeuratObject`.\

2.  Assess the correlation between metrics included in `meta.data`
    generating a pairsplot.\

3.  Create violin plots resembling QC metrics using `Seurat` function
    `VlnPlot()`.\

4.  Visualize QC features over the HE image exploiting
    `SpatialDimPlot()` and `SpatialFeaturePlot()`.\

5.  Remove spots with very few transcript/gene counts.\
    By applying permissive thresholds, we exclude cells with
    insufficient gene expression information, which cannot be used for
    downstream analysis.\
    In particular, consider these two thresholds:\
    -   `nFeature_threshold`: compute the 25th percentile of `nFeature`
    distribution\
    -   `nCount_threshold`: compute the 25th percentile of `nCount`
    distribution\
\

6. Select spots having: \
    - `nCount` \> `nCount_threshold` \
    - `nFeature` \> `nFeature_threshold`




```{=html}
  <hr>
```


**Questions** 

A) Inspecting the QCs, how do you evaluate this experiment? \


B) How many spots did you retained after applying filtering strategy?\



```{=html}
  <hr>
```


## Answer {.unnumbered}


```{r}
# Your code here













```


```{=html}
  <hr>
```



# Identification of cluster of spots {.tabset .tabset-fade .tabset-pills}



## Question {.unnumbered}

Let's define clusters of spots that share similar transcriptomic
profiles.

1.   Normalize counts using log-normalization method and find the top
    2,000 variable features using `FindVariableFeatures()`.

2.   Perform PCA dimensionality reduction computing first 50 PCs and
    scale data.

3.   Find the 20-nearest neighbors of each spot using the function
    `FindNeighbors()`.

4.   Identify clusters using `PCA reduction` as reduction for the
    nearest-neighbor graph construction. Specifically, use the first 30
    dimensions of PCA as input.\

5.   Visualize the identified clusters over the HE image using
    `SpatialDimPlot()` function labelling each cluster with the
    correspondent number.

6.   Visualize the identified clusters over the HE image using `SpatialDimPlot()` function labelling each cluster with the correspondent number.


7.  Collect in a list the spot names according to the correspondent cluster using the function `CellsByIdentities()`.\
    Plot the clusters separately over the HE image specifying in
    `cells.highlight` parameter the list containing groups of cells per cluster.


```{=html}
  <hr>
```


**Questions**

A.  Which parameter in the function `FindClusters()` impacts the most on the numbers of detected clusters? You can set this parameter equal to 3 and see if you get a larger number of communities.\

B.  How many clusters did you find? \

C.  How many spots did you get in each cluster?

D.  What algorithm did you choose to optimize modularity? Motivate your choice.



```{=html}
  <hr>
```



## Answer {.unnumbered}


```{r}
# Your code here















```



```{=html}
  <hr>
```




# Getting the gene signatures {.tabset .tabset-fade .tabset-pills}


## Question {.unnumbered}

1.  Use the function `FindAllMarkers()` to get the list of genes that
    represent each cluster keeping only positive markers (FoldChange (FC) > 0).\
    In the function, choose proper thresholds for parameters in order to
    limit testing procedures: \

    -   to genes having at least X-fold difference (log-scale)\
    -   to consider only genes detected in a minimum fraction of X cells in clusters\

2.  Rank markers per cluster according to the adjusted p-value.


3.  Subset in another variable differentially expressed markers having a `log2(fold-change) ≥ 1` and `adjusted p-value ≤ 0.1`.\


4.  Visualize espression of top 50 markers per cluster in a heatmap.





```{=html}
  <hr>
```


**Questions**

A)   What are the parameters that you set in `FindAllMarkers()`? What statistical test did you choose to identify markers ? 


B)   How many positive markers did you identify per cluster? Of those,
    how many are differentially expressed ?


C)   In Seurat adjusted p-values are calculated using what type of multiple testing correction algorithm?




```{=html}
  <hr>
```


## Answer {.unnumbered}


```{r}
# Your code here















```


```{=html}
  <hr>
```



# Functional analysis on markers {.tabset .tabset-fade .tabset-pills}

Let's identify cancerous and not cancerous cells by annotating markers
on Hallmarks of cancer and cancer/healthy drivers retrieved from Network
of Cancer Genes and Healthy Drivers.\

To do so, we will exploit:

-   the involvement of genes in the Hallmarks of Cancer database\
-   the selective advantage of cancer and healthy drivers genes\

In this phase, consider all positive markers you identified per cluster.
In this way, you will get a comprehensive view of the genes composing each
cluster.




```{=html}
  <hr>
```


## Question {.unnumbered}

### **Hallmarks of cancer**


To investigate Hallmarks of Cancer features :\


  1.  Retrieve human Hallmarks gene set from [MSigDB Collections](https://www.gsea-msigdb.org/gsea/msigdb/collections.jsp) using the `msigdbr()` function by specifying the abbreviation category correspondent to Hallmarks collection.\


  2.  Perform an enrichment analysis with the `enricher()` function on all positive markers identified for each cluster.\
    Employ the following cutoffs: `pvalue = 1` and `qvalue = 1`. For multiple testing correction method, use the `Benjamini–Hochberg method`.\
    Select statistically significant enrichments (FDR \< 0.1).\


  3.  Generate a plot visualizing enriched hallmark-related pathways on y-axis, -log10(FDR) on the x-axis and color by cluster identifiers.


### **Network of Cancer Genes & Healthy Drivers**

Now:\

1.  Download the manually curated collection of cancer genes and healthy
    drivers from [Network of Cancer Genes & Healthy Drivers (NCG) portal](http://network-cancer-genes.org/). Alternatively, you can
    find the already downloaded data in the `input` folder.\

    | What                        | Path to                                                          |
    |:------------------|:----------------------------------------------------|
    | Drivers of cancer clone     | `../input/NCG_cancerdrivers_annotation_supporting_evidence.tsv/` |
    | Drivers of not cancer clone | `../input/NCG_healthydrivers_annotation_supporting_evidence.tsv` |

\

2.  Subset the cancer driver genes primarily associated with ovarian cancer and
    healthy driver genes associated with the gynecologic organ system.\


3.  In the dataframe containing all positive markers for each cluster, add
    two columns to annotate whether each gene is a cancer driver or 
    a healty driver according to NCG collection.\


4.  Sum up the number of `drivers of cancer clone` and
    `drivers of not cancer clone` expressed in each spot.\
    Add these annotations to the meta.data table of your `SeuratObject`.
    
    
5.  Visualize these two newly added features over the HE image of the
    tissue.




```{=html}
  <hr>
```



### **Questions**

A)  Considering enrichment of clusters in Hallmark pathways, what is the most enriched gene set in each cluster?

      *HINT: you can explore metrics such as RichFactor or GeneRatio.*


B)  By inspecting the enriched hallmarks of cancer,
    can you assign a specific label to each cluster amongts the following list?\
    - `normal`\
    - `early tumor onset`\
    - `late tumor stage`\
    - `inflammation`\
    - `DNA damage`\


C)  How many cancer genes and healthy genes we are dealing with?


D)  Looking at the HE image, are clusters expressing drivers of cancer or not cancer clone
    clearly separated within the tissue?


E)  Based on the expression of cancer drivers, where would you expect to 
    find cancer lesions in the tissue? Take the HE image of the tissue and
    draw a box where you expected to localize tumor cells in the tissue.


F)  In which Hallmark pathways are the clusters that express cancer driver genes enriched?
    And in those featured by drivers of not cancer clones?



```{=html}
  <hr>
```

## Answer {.unnumbered}


```{r}
# Your code here














```


```{=html}
  <hr>
```


# Prediction cellular types by unsupervised approach {.tabset .tabset-fade .tabset-pills}

Let's see whether we can strengthen the results by inferring cell
composition.\
Cell type prediction can be done using an unsupervised approach based on
[`clustermole` package](https://igordot.github.io/clustermole/).\
Clustermole performs cell type prediction based on a set of marker genes
or on a table of expression values. We will infer cell types from gene
expression matrix.



```{=html}
  <hr>
```



## Question {.unnumbered}

1)  From SeuratObject retrieve assay data containing normalized counts
    and transform it into a matrix.\


2)  Explore the `clustermole` package to find the function to perform cell
    type composition inference based on normalized expression counts.\
    Among available enrichment methods use `ssGSEA`, as it is considered to be
    one of most accurate methods for gene set enrichment (GSE) analyses
    *(Lauria A., et al., NAR, 2020)* .\


3)  Retrieve the full list of human cell type markers in the clustermole
    database using `clustermole_markers()`.\
    Subset the list to include only rows matching the patterns `Human`
    and `CellMarker` in the `celltype_full` column.\
    From this selection, further filter markers that meet **at least one**
    of the following conditions:\
    - related to `ovary` as organ\
    - **not** containing one of the following patterns:
    `|Pluripotent|Pancrea|Hepat|Sertoli|Oligo|Induced|Hematopoietic|Plasma|Mast|Pluripotentstemcell|Platelet|Megakaryocyte|Embr|Neur|Glia|glia|Purkinje|Pyrami|Germ|germ|Follic|neuro|Adventitial`.


4)  Subset the cell type enrichments obtained in step 2, keeping only those cell
    types selected in the previous step.\
    For each spot, select the most enriched cell type according to the
    `score_rank.` Now, you can add cell type prediction per spot in the
    `meta.data` table of your `SeuratObject`.\


5)  In the cell type enrichment dataframe add a column that associates 
    each spot barcode name with its corresponding cluster number ID.
    Calculate the proportion of each cell type within each cluster.\
    Generate a barplot plotting on x-axis the different cluster and
    filled by the proportion of cell types.\





```{=html}
  <hr>
```



**Questions**\


A)   Considering results in the barplot, which clusters contain an higher fraction of
    cancer cells?


B)   Calculate cell types per cluster and spatially visualize them on the tissue
    image. Which cell types compose clusters expressing cancer drivers
    measured in the previous section?


C)   Is there any concordance between top expressed markers and inferred cell
    types per cluster?
   
    
D)   What are the limitations of clustermole in the context of 10X Visium data? 



```{=html}
  <hr>
```


## Answer {.unnumbered}

```{r}
# Your code here















```


```{=html}
  <hr>
```



# Identification of ligand receptors (LR) interactions {.tabset .tabset-fade .tabset-pills}

## Question {.unnumbered}

1.  Detect LR interactions of **each cell type** using
    [LIANA](https://www.nature.com/articles/s41467-022-30755-0).\
    Specifically, select [natmi](https://github.com/forrest-lab/NATMI), [cell_italk](https://github.com/Coolgenome/iTALK), [call_cellchat](https://github.com/jinworks/CellChat) and [sca](https://github.com/SCA-IRCM/SingleCellSignalR) as methods and [OmniPath](https://github.com/saezlab/OmnipathR) as resource.\
    Ensure that a minimum of **5** cells per cell type is considered for the analysis.\
    Include these additional columns in your results: `ligand.expr`,
    `receptor.expr`, `ligand.pval`, `receptor.pval`, `ligand.FDR`,
    `receptor.FDR`.\


2.  Combine the results and specify the ranking of each interaction
    according the `aggregate_rank` score.\


3.  Explore visualization function provided by LIANA. Highlight LR
    interactions involving cancer cells.
    
    
4. Assess ligand-receptors interactions among **clusters**. Do you observe concordant results in terms of LR pairs characterizing tumoral clusters?  





```{=html}
  <hr>
```



**Questions**

A)  Which are the top-5 ranked LR interactions per cell type?


B)  Which are LR interactions involving cancer cells as target? Is there a most recurrent interactor?




```{=html}
  <hr>
```



## Answer {.unnumbered}


```{r}
# Your code here















```

```{=html}
  <hr>
```


